---
title: "Epigenomics Roadmap Hands-on"
author: "Jan Izquierdo"
date: "2024-05-24"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Data preparation
```{r importing and preparing, warning=FALSE}
library(dplyr)
source("/home/jj/bedtools2/BEDtoolsR.R")

gen_path<-"/home/jj/Desktop/Bioinformatics/2nd_year/3term/"
fullpath<-paste0(gen_path,"Omics_Techniques/Seminars/3.1-Epigenomics_roadmap/metadata.roadmap_clean.txt")

metadata<-read.table(fullpath ,sep="\t")
#View(metadata)

#Distinguish Fetal and Non-fetal(Adult) samples
metadata$PERIOD<-ifelse(grepl("fetal|Fetal", metadata$V3), "Fetal", "Adult")

#Create a new column with UniqueNames incorporating the tissue, PERIOD and ID information
metadata$UniqueName<-paste0(metadata$V2, "_", metadata$PERIOD, "_", metadata$V1)

#Choosing tissues
tissues<-c("Brain", "brain", "Muscle", "muscle", "Digestive", "digestive", "Heart", "heart") 

filtered_metadata<-dplyr::filter(metadata, V2 %in% tissues)$V1 #Teacher's method

filtered_metadata<-metadata[grepl(paste(tissues, collapse="|"), metadata$V2)==TRUE,] #database of only 
#the files that interest us (all that contain any of the words of tissues in column metadata$V2)
```

Reading files and storing them in a list
```{r file content to a list, message=FALSE, results=FALSE}
roadmap<-list()

for (f in filtered_metadata$V1){ #first column contains the id that differs the files from each other 
                                 #(id of entry as well)
  filename<-paste0("all.mnemonics.bedFiles/",f,"_18_core_K27ac_mnemonics.bed.gz") #f is the ids
  if (file.exists(filename)){ #in case that a file does not exist, avoid errors
    print(f)
    roadmap[[f]]<-read.table(gzfile(filename)) #all contents of file to roadmap 
                                               #"dictionary" with the ID as key
  }
}

#Changing ID "keys" to UniqueNames that we generated before
library(plyr)
names(roadmap)<-mapvalues(names(roadmap), from=metadata$V1, to=metadata$UniqueName)
```
```{r part 2 to be able to see results}
names(roadmap)

roadmap[["Muscle_Fetal_E089"]] %>% head()

#avoid executing this chink every time by saving roadmap
#saveRDS(roadmap, "roadmap.rds")
```

## 1.Calculate pairwise Jaccard index

Prepare the data and create the matrix
```{r 1.preparation}
#Load roadmap if the previous chunk hasn't been executed
roadmap<-readRDS("roadmap.rds")

#Doing the intersection
state="1_TssA"
b1<-dplyr::filter(roadmap[["Muscle_Fetal_E090"]], V4==state) #V4 are the promoters 
#(if we did all genomes it'd almost the same), we choose 1 promoter
b2<-dplyr::filter(roadmap[["Digestive_Adult_E102"]], V4==state)

bedTools.2jac(bed1=b1,bed2=b2)

#do for all members in names(roadmap) and place in a matrix
rnames<-names(roadmap)
m<-matrix(nrow=length(rnames), ncol=length(rnames), dimnames=list(rnames)) #matrix of rnames dimensions
colnames(m)<-rnames
```

Fill the jaccard matrix
```{r 1.fill and save the matrix, results=FALSE, warning=FALSE}

for (i in 1:length(rnames)){
  for (j in 1:length(rnames)){
    a<-dplyr::filter(roadmap[[i]], V4==state)
    b<-dplyr::filter(roadmap[[j]], V4==state)
    index_num<-bedTools.2jac(bed1=a,bed2=b) #MYFUNCTION #2jac for testing
    colnames(index_num)<-make.names(index_num[1,])
    index_num<-index_num[2,]
    m[i,j]<-index_num$jaccard #store jac2 indexes there #or intersection?
  }
}
saveRDS(m, "jaccard_matrix.rds")
```

Load the jaccard matrix
```{r 1.load the matrix}
m<-readRDS("jaccard_matrix.rds")
```

## 2.Visualize Jaccard Index matrix in a heatmap, indicating tissue of origin and PERIOD

Prepare the data for the heatmap and create it
```{r 2. m in heatmap, warning=FALSE}
library(pheatmap)

annotation_col <- data.frame( Tissue = filtered_metadata$V2, Period = filtered_metadata$PERIOD, 
                              row.names = filtered_metadata$UniqueName)
m <- as.matrix(m)
m <- apply(m, 2, as.numeric)

pheatmap(m,
  annotation_col = annotation_col,
  main = "Heatmap of Jaccard Indexes",
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  show_rownames = FALSE,
)
```

## 3.Report jaccard index between sample E071 and sample E074

Set up variables and consult the index
```{r 3.define variables for each}
#m<-readRDS("jaccard_matrix.rds")
#o bé
rownames(m)<-rnames

#E071 on filtered -> Name in UniqueNames -> m[name1] is row of name -> m[name1, name2] 
#m[name1, name2] is row and column of names
E071_Unique<-filtered_metadata$UniqueName[filtered_metadata$V1=="E071"]
E074_Unique<-filtered_metadata$UniqueName[filtered_metadata$V1=="E074"]

m[E071_Unique, E074_Unique]
```

## 4.Perform muldimensional scaling on a distance matrix based on 1-jaccardIndex.

Creation of the 1-jaccard distance matrix and performing the multidimensional scaling
```{r 4.creating the distance matrix}
#distance matrix is 1-jaccard distance
dm<-1-m

#Perform the multimensional scaling, k=4 for 4 dimensions ( using 4 for the next exercise )
multi_d_scaling <- cmdscale(dm, k = 4)

```

## 5.Plot 1:2 and 3:4 dimensions and color by tissue, shape by fetal/adult.

Create a new dataframe for plotting the dimensions
```{r 5.create a new dataframe to plot the dimensions}
#rename the coluns from [,1] to their corresponding dimension and transorm it into a dataframe
colnames(multi_d_scaling)<-c("MDS1", "MDS2", "MDS3", "MDS4")
df_multi_d_scaling <- as.data.frame(multi_d_scaling)

#Create a column with the rownames,for easier join with filtered_metadata(needed for PERIOD and tissue)
df_multi_d_scaling$Name<-rownames(df_multi_d_scaling)
df_multi_d_scaling <- df_multi_d_scaling %>%
  left_join(filtered_metadata, by = c("Name" = "UniqueName"))
```

Plot for dimensions 1 and 2
```{r 5.Plot dimensions 1 and 2}
library(ggplot2)
ggplot(df_multi_d_scaling, aes(x = MDS1, y = MDS2, color = V2, shape = PERIOD))+geom_point(size = 5)+
  labs(x="Dimension 1", y="Dimension 2")
```

Plot for dimensions 3 and 4
```{r 5.Plot dimensions 3 and 4}
ggplot(df_multi_d_scaling, aes(x = MDS3, y = MDS4, color = V2, shape = PERIOD))+geom_point(size = 5)+
  labs(x="Dimension 3", y="Dimension 4")
```

## 6.Compute hierarchical clustering in jaccard distance matrix with R function “hclust”, cut the dendrogram at different Ks using R function “cutree” and plot the dendrogram coloring the obtained clusters of samples. (You can use the R Package dendextend)

Prepare the data and compute the clustering
```{r 6.Data preparation and hclus execution}
library(dendextend)

dmatrix_prep<-as.dist(dm)
computed_clusters<-hclust(dmatrix_prep, method="complete")

plot(computed_clusters, sub="", xlab="", cex=0.9)
```

Cluster separation and plotting with cutree
```{r 6.Unsing cutree to cut and represent the data}

cluster_numbers<-c(2, 3, 4, 6, 8)

#par(mfrow = c(1, length(cluster_numbers)/2))
for (cl in cluster_numbers){
  clusters <- cutree(computed_clusters, k = cl)
  dend <- color_branches(as.dendrogram(computed_clusters), k = cl)
  par(mar = c(10, 4, 2, 2)+0.1)
  plot(dend, main=paste0("Diagram of ", cl, " clusters"), cex=0.9)
}

```

### Do you see a logic in the clusters using epigenomes?

```{r 6.Discussion 1, echo=FALSE}
cat("Most clusters seem to be separated by tissue and the separation occurs by earliest branch 
division (On 2 clusters, it divides by red and blue instead of creating a division out of 
green, which is the new cluster created in the 3 clusters plot)\n")
```

### Which major sub-divisions can you see?

```{r 6.Discussion 2, echo=FALSE}
cat("The main subdivisons on 3 clusters are Muscle+Small muscle, Brain and Digestive+Heart, which 
we can observe that also get separated when we increase the number of clusters in later plots\n")
```

### All states convey similar information?

```{r 6.Discussion 3, echo=FALSE}
#NO HO SE
cat("The further away 2 samples are, the more epigenomic differences they present.\n")
```